Hierarchy Builder: algebraic~hierarchies made easy in Coq with Elpi

### 2020-05-04 hb.tex
Update hb.tex
Fix some typesetting issues

\newcommand{\Lang}{\ensuremath{\Lambda}}
\newcommand{\term}{term}
\newcommand{\mixin}{mixin}
\newcommand{\mixins}{mixins}
\newcommand{\Mixins}{Mixins}
\newcommand{\M}{\ensuremath{\mathcal{M}}}
\newcommand{\factory}{factory}
\newcommand{\factories}{factories}
\newcommand{\Factories}{Factories}
% \newcommand{\packager}{packager}
% \newcommand{\packagers}{packagers}
% \newcommand{\Packager}{Packager}
\newcommand{\phantterm}{abbreviation}
\newcommand{\phantterms}{abbreviations}
\newcommand{\Phantterms}{abbreviations}
\newcommand{\builder}{builder}
\newcommand{\mixinbuilder}{builder}
\newcommand{\mixinbuilders}{builders}
\newcommand{\factoryinstance}{factory instance}
\newcommand{\factoryinstances}{factory instances}
\newcommand{\F}{\ensuremath{\mathcal{F}}}
\newcommand{\dep}{\ensuremath{\mathit{dep}}}
\newcommand{\powerset}[1]{\raisebox{.15\baselineskip}{\Large\ensuremath{\wp}}(#1)}
\newcommand{\C}{\ensuremath{\mathcal{C}}}
\newcommand{\B}{\ensuremath{\mathcal{B}}}
\newcommand{\class}{class}
\newcommand{\classes}{classes}
\newcommand{\cdef}{\ensuremath{\mathit{def}}}
\newcommand{\Str}{\ensuremath{\mathcal{S}}}
\newcommand{\structure}{structure}
\newcommand{\structures}{structures}
\newcommand{\issubclass}{\ensuremath{\leq}}
\newcommand{\subclass}{subclass}
\newcommand{\join}{\ensuremath{\mathit{join}}}
\newcommand{\requires}{\ensuremath{\mathit{requires}}}
\newcommand{\target}{\ensuremath{\mathit{target}}}
\newcommand{\provides}{\ensuremath{\mathit{provides}}}
\newcommand{\from}{\ensuremath{\mathsf{from}}}
\newcommand{\proj}[2]{\ensuremath{#1[#2]}}
\newcommand{\set}[1]{\left\{#1\right\}}
\newcommand{\enum}[2]{\ensuremath{\set{#1,\ldots,#2}}}
\newcommand{\vect}[1]{\overrightarrow{#1}}
\newcommand{\pmp}[1]{\ensuremath{\vect{\left(p : m\ T\ p_\sigma\right)}^{#1}}}
\newcommand{\hbmixin}{{\tt\color{dkgreen}HB.mixin}}
\newcommand{\hbfactory}{{\tt\color{dkgreen}HB.factory}}
\newcommand{\hbinstance}{{\tt\color{dkgreen}HB.instance}}
\newcommand{\hbstructure}{{\tt\color{dkgreen}HB.structure}}
\newcommand{\hbbuilders}{{\tt\color{dkgreen}HB.builders}}
\newcommand{\hbend}{{\tt\color{dkgreen}HB.end}}

@@

From now on the axioms as well as the abstract theory of rings apply to
integers, as shown in lemma \coq{exercise}. The details of the proof
do not matter here, what is worth pointing out is that in a single
statement we mix monoid (e.g. \coq{+}) and ring (e.g. \coq{-}) operations and in the
proof we use monoid axioms (e.g. \coq{addrA}), ring axioms
(e.g. \coq{addrN}) and ring lemmas (e.g. \coq{addrC}), all seamlessly.

@@

\begin{definition}[\M{}, \mixins{}]
  A \emph{\mixin{}} \(m \in \M{}\) is a distinguished Coq record with one or more
  parameters.  The first parameter must be a \coq{(T : Type)}, while
  the other parameters are \mixins{} \(\left(m_i\right)_{i \in \enum1n}\), each of which
  is applied to \coq{T} and possibly previous mixin variables. I.e.
\begin{coqcode}
Record #$m$# (T : Type) #\(\pmp{n}\)# : Type := { .. }.
\end{coqcode}
where \(\pmp{n} = (p_1 : m_1\ T) \ldots \left(p_n : m_n\ T\ p_{\sigma(n,1)}\ \ldots\ p_{\sigma(n,q_n)}\right)\)
and where for all \(i~\in~\enum1n\) we have \(q_i \in \mathbb{N}\) and the
arguments of \(m_i \in \M{}\) consist in $q_i$ of the previously quantified
mixin parameters $p_k$, i.e. for all \(k \in \enum1{q_i}\), we
have \(\sigma(i,k) \in \enum1{i-1}\).
\end{definition}

\begin{definition}[\dep{}, \mixin{} dependencies]
  Given \(m \in \M{}\), we define \(\dep{}(m) \in \powerset\M{}\)
  as the set of all \mixins{} that occur as parameters of \(m\), i.e.
  \(\dep{}(m) = \enum{m_1}{m_n}\).
\end{definition}

@@

\begin{definition}[$\requires{}$, factory requirements.]
  Given \(f \in \F{}\), we define \(\requires{}(f) \in \powerset\M{}\)
  as the set of all \mixins{} that occur as parameters of \(f\), i.e.
  \(\dep{}(f) = \enum{m_1}{m_n}\).
\end{definition}

The following property holds,
\begin{proposition}[\requires{} is closed under \dep{}]
\(\forall f \in \F{}, \dep{}^\star(\requires{}(f)) \subseteq \requires{}(f)\)
\end{proposition}
\begin{proof}
Because Coq records and global definitions are well typed in the empty context.
\end{proof}

@@

\begin{definition}[$\provides{}$]
  \( \provides{}(f) = \{m | \from{}(f,m) \textrm{ is defined}\}
    \in \powerset\M{}\) the set of mixins that a factory
    \(f \in \F{}\) provides, through its \mixinbuilders{}.
\end{definition}

@@

\begin{proposition}[\( \M{} \subseteq \F{} \)] There is a way to see mixins as factories.
\end{proposition}
\begin{proof}
  For all \(m \in \M{}\) we have \(\requires{}(m) = \dep{}(m)\),
  \(\provides{}(m) = \{m\}\) and
  \(\from{}(m,m) = \left(\coq{fun T }\ \pmp{n} \ \coq{(x :}\ m\
    \coq{T }\ p_1\ \ldots p_n\ \coq{) => x}\right) \in \B{}.\)
\end{proof}

@@

\begin{definition}[\cdef{}, class definition]
  We call \(\cdef{}(c) \in \powerset\M{}\) the set of \mixins{}
  mentioned in the fields of the \class{}, i.e.
  \(\{m_1, \ldots,m_n\}\).  Given that \class{} records are well typed
  in the empty context the set of \mixin{} records is closed
  transitively.  The implementation enforces that no two \class{}
  records contains the same set of \mixins{} (disregarding the order of
  the fields), i.e. \cdef{} is injective.
\end{definition}

@@

Classes could also be seen as factories.
\begin{proposition}[\( \C{} \subseteq \F{} \)]
  forall \(m \in \cdef{}(c)\) we have \(\from{}(c,m_i) = p_i,\) and it
  follows that \(\requires{}(c)=\emptyset \),
  \(\provides{}(c)=\cdef{}(c).\)
\end{proposition}
However since classes are generated, their implementation may change,
thus users should not rely on constructors of classes. Hence the only
way users may refer to a class is as an argument of \hbmixin{},
\hbfactory{} or \hbstructure{}.

@@

Since \mixins{} may change but \factories{} stay the same, \HB{}
commands must never rely on a particular set of \mixins{} as
arguments, and \factory{} arguments must never be given explicitly by
the user. As described in Sections~\ref{sec:hb-mixins-factories}
and~\ref{sec:hb-class-struct}, \HB{} commands take a list of
\factories{} as arguments, which they expand into lists of \mixins{}
behind the scene. However \factory{} types and constructors have
\mixin{} arguments that must be inferred automatically when used. To
this end, the commands \hbmixin{} and \hbfactory{} generate
\phantterms{} for the user to replace uses of constructors of \factories{}.
These commands first create a record \(f_{aux}\) with a
constructor \(F_{aux}\) and then create abbreviations \(f\) and \(F\)
that automatically fill the mixin arguments of \(f_{aux}\) and \(F_{aux}\)
respectively.
See \cite[Section 7]{DBLP:conf/itp/MahboubiT13} for a detailed description
of how to implement these abbreviations in Coq.

\begin{coqcode}
Record #\(f_{aux}\)# T #\(\pmp{n}\)# := #\(F_{aux}\)# { ... }.
Notation #$f$# T := (#\(f_{aux}\)# T ...(* $p_1 \ldots p_n$ inferred when $T$ is known *)).
Notation #$F$# T  #\(x_1\ \ldots\ x_k\)# := (#\(F_{aux}\)# T ...(* $p_1 \ldots p_n$ inferred when $T$ is known *) #\(x_1\ \ldots\ x_k\)#).
Definition b : #$f$# T := #$F$# T #\(x_1\ \ldots\ x_n\)#
\end{coqcode}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Conclusion}
In this paper we design and implement \HB{}, a high level language to describe
hierarchies of algebraic structures. The implementation of \HB{} is
based on the Elpi extension language for the Coq system and is available
at \url{https://github.com/math-comp/hierarchy-builder}.
The implementation amounts to approximately a thousand lines of (commented)
Elpi code and tree hundred lines of Coq vernacular. It took less than one month
to implement \HB{} but it took several years of attempts and fruitful
discussions with the people we thank below.

>>>>

\newcommand{\Lang}{\ensuremath{\Lambda}}
\newcommand{\term}{term}
\newcommand{\mixin}{mixin}
\newcommand{\mixins}{mixins}
\newcommand{\Mixins}{Mixins}
\newcommand{\M}{\ensuremath{\mathcal{M}}}
\newcommand{\factory}{factory}
\newcommand{\factories}{factories}
\newcommand{\Factories}{Factories}
% \newcommand{\packager}{packager}
% \newcommand{\packagers}{packagers}
% \newcommand{\Packager}{Packager}
\newcommand{\phantterm}{abbreviation}
\newcommand{\phantterms}{abbreviations}
\newcommand{\Phantterms}{abbreviations}
\newcommand{\builder}{builder}
\newcommand{\mixinbuilder}{builder}
\newcommand{\mixinbuilders}{builders}
\newcommand{\factoryinstance}{factory instance}
\newcommand{\factoryinstances}{factory instances}
\newcommand{\F}{\ensuremath{\mathcal{F}}}
\newcommand{\dep}{\ensuremath{\mathit{dep}}}
\newcommand{\powerset}[1]{\raisebox{.15\baselineskip}{\Large\ensuremath{\wp}}(#1)}
\newcommand{\C}{\ensuremath{\mathcal{C}}}
\newcommand{\B}{\ensuremath{\mathcal{B}}}
\newcommand{\class}{class}
\newcommand{\classes}{classes}
\newcommand{\cdef}{\ensuremath{\mathit{def}}}
\newcommand{\Str}{\ensuremath{\mathcal{S}}}
\newcommand{\structure}{structure}
\newcommand{\structures}{structures}
\newcommand{\issubclass}{\ensuremath{\leq}}
\newcommand{\subclass}{subclass}
\newcommand{\join}{\ensuremath{\mathit{join}}}
\newcommand{\requires}{\ensuremath{\mathit{requires}}}
\newcommand{\target}{\ensuremath{\mathit{target}}}
\newcommand{\provides}{\ensuremath{\mathit{provides}}}
\newcommand{\from}{\ensuremath{\mathsf{from}}}
\newcommand{\proj}[2]{\ensuremath{#1[#2]}}
\newcommand{\set}[1]{\left\{#1\right\}}
\newcommand{\enum}[2]{\ensuremath{\set{#1,\ldots,#2}}}
\newcommand{\vect}[1]{\overrightarrow{#1}}
\newcommand{\pmp}[1]{\ensuremath{\vect{\left(p : m\ T\ p_\sigma\right)}^{#1}}}
\newcommand{\aux}{\mathit{aux}}
\newcommand{\hbmixin}{{\tt\color{dkgreen}HB.mixin}}
\newcommand{\hbfactory}{{\tt\color{dkgreen}HB.factory}}
\newcommand{\hbinstance}{{\tt\color{dkgreen}HB.instance}}
\newcommand{\hbstructure}{{\tt\color{dkgreen}HB.structure}}
\newcommand{\hbbuilders}{{\tt\color{dkgreen}HB.builders}}
\newcommand{\hbend}{{\tt\color{dkgreen}HB.end}}

@@

From now on the axioms as well as the abstract theory of rings apply to
integers, as shown in lemma \coq{exercise}. The details of the proof
do not matter here, what is worth pointing out is that in a single
statement we mix monoid (e.g.~\coq{+}) and ring (e.g.~\coq{-}) operations and in the
proof we use monoid axioms (e.g.~\coq{addrA}), ring axioms
(e.g.~\coq{addrN}) and ring lemmas (e.g.~\coq{addrC}), all seamlessly.

@@

\begin{definition}[\M{}, \mixins{}]
  A \emph{\mixin{}} \(m \in \M{}\) is a distinguished Coq record with one or more
  parameters.  The first parameter must be a \coq{(T : Type)}, while
  the other parameters are \mixins{} \(\left(m_i\right)_{i \in \enum1n}\), each of which
  is applied to \coq{T} and possibly previous mixin variables. I.e.
\begin{coqcode}
Record #$m$# (T : Type) #\(\pmp{n}\)# : Type := { .. }.
\end{coqcode}
where \(\pmp{n} = (p_1 : m_1\ T) \ldots \left(p_n : m_n\ T\ p_{\sigma(n,1)}\ \ldots\ p_{\sigma(n,q_n)}\right)\)
and where for all \(i~\in~\enum1n\) we have \(q_i \in \mathbb{N}\) and the
arguments of \(m_i \in \M{}\) consist in $q_i$ of the previously quantified
mixin parameters $p_k$, i.e.~for all \(k \in \enum1{q_i}\), we
have \(\sigma(i,k) \in \enum1{i-1}\).
\end{definition}

\begin{definition}[\dep{}, \mixin{} dependencies]
  Given \(m \in \M{}\), we define \(\dep{}(m) \in \powerset\M{}\)
  as the set of all \mixins{} that occur as parameters of \(m\),
  i.e.~\(\dep{}(m) = \enum{m_1}{m_n}\).
\end{definition}

@@

\begin{definition}[$\requires{}$, factory requirements.]
  Given \(f \in \F{}\), we define \(\requires{}(f) \in \powerset\M{}\)
  as the set of all \mixins{} that occur as parameters of \(f\),
  i.e.~\(\dep{}(f) = \enum{m_1}{m_n}\).
\end{definition}

The following property holds,
\begin{proposition}[\requires{} is closed under \dep{}]
\(\forall f \in \F{}, \dep{}^\star(\requires{}(f)) \subseteq \requires{}(f).\)
\end{proposition}
\begin{proof}
Because Coq records and global definitions are well typed in the empty context.
\end{proof}

@@

\begin{definition}[$\provides{}$]
  \( \provides{}(f) = \{m \mid \from{}(f,m) \textrm{ is defined}\}
    \in \powerset\M{}\) the set of mixins that a factory
    \(f \in \F{}\) provides, through its \mixinbuilders{}.
\end{definition}

@@

\begin{proposition}[\( \M{} \subseteq \F{} \)] There is a way to see mixins as factories.
\end{proposition}
\begin{proof}
  For all \(m \in \M{}\) we have \(\requires{}(m) = \dep{}(m)\),
  \(\provides{}(m) = \{m\}\) and
  \(\from{}(m,m) = \left(\coq{fun T }\ \pmp{n} \ \coq{(x :}\ m\
    \coq{T }\ p_1\ \ldots p_n\coq{) => x}\right) \in \B{}.\)
\end{proof}

@@

\begin{definition}[\cdef{}, class definition]
  We call \(\cdef{}(c) \in \powerset\M{}\) the set of \mixins{}
  mentioned in the fields of the \class{}, i.e.~\(\{m_1, \ldots,m_n\}\).
  Given that \class{} records are well typed
  in the empty context the set of \mixin{} records is closed
  transitively.  The implementation enforces that no two \class{}
  records contains the same set of \mixins{} (disregarding the order of
  the fields), i.e.~\cdef{} is injective.
\end{definition}

@@

Classes could also be seen as factories.
\begin{proposition}[\( \C{} \subseteq \F{} \)]
  For all \(m \in \cdef{}(c)\) we have \(\from{}(c,m_i) = p_i,\) and it
  follows that \(\requires{}(c)=\emptyset \),
  \(\provides{}(c)=\cdef{}(c).\)
\end{proposition}
However since classes are generated, their implementation may change,
thus users should not rely on constructors of classes. Hence the only
way users may refer to a class is as an argument of \hbmixin{},
\hbfactory{} or \hbstructure{}.

@@

Since \mixins{} may change but \factories{} stay the same, \HB{}
commands must never rely on a particular set of \mixins{} as
arguments, and \factory{} arguments must never be given explicitly by
the user. As described in Sections~\ref{sec:hb-mixins-factories}
and~\ref{sec:hb-class-struct}, \HB{} commands take a list of
\factories{} as arguments, which they expand into lists of \mixins{}
behind the scene. However \factory{} types and constructors have
\mixin{} arguments that must be inferred automatically when used. To
this end, the commands \hbmixin{} and \hbfactory{} generate
\phantterms{} for the user to replace uses of constructors of \factories{}.
These commands first create a record \(f_{\aux}\) with a
constructor \(F_{\aux}\) and then create abbreviations \(f\) and \(F\)
that automatically fill the mixin arguments of \(f_{\aux}\) and \(F_{\aux}\)
respectively.
See \cite[Section 7]{DBLP:conf/itp/MahboubiT13} for a detailed description
of how to implement these abbreviations in Coq.

\begin{coqcode}
Record #\(f_{\aux}\)# T #\(\pmp{n}\)# := #\(F_{\aux}\)# { ... }.
Notation #$f$# T := (#\(f_{\aux}\)# T ...(* $p_1 \ldots p_n$ inferred when $T$ is known *)).
Notation #$F$# T  #\(x_1\ \ldots\ x_k\)# := (#\(F_{\aux}\)# T ...(* $p_1 \ldots p_n$ inferred when $T$ is known *) #\(x_1\ \ldots\ x_k\)#).
Definition b : #$f$# T := #$F$# T #\(x_1\ \ldots\ x_n\)#.
\end{coqcode}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Conclusion}
In this paper we design and implement \HB{}, a high level language to describe
hierarchies of algebraic structures. The implementation of \HB{} is
based on the Elpi extension language for the Coq system and is available
at \url{https://github.com/math-comp/hierarchy-builder}.
The implementation amounts to approximately a thousand lines of (commented)
Elpi code and tree hundred lines of Coq vernacular. It took less than one month
to implement \HB{}, but its design took several years of attempts and fruitful
discussions with the people we thank below.
